---
title: "Part 1 Markdown"
author: "Felix,Zhikai"
date: "2024-04-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part 1
## a) Apply the random walk Metropolis algorithm using N = 10000 and s = 1. Use the generated samples (x1, . . . xN ) to construct a histogram and a kernel density plot in the same figure. Note that these provide estimates of f (x).Overlay a graph of f (x) on this figure to visualise the quality of these estimates. Also, report the sample mean and standard deviation of the generated samples.

```{r}
library(tidyverse)
```

The probability density function is defined
```{r}
pdf_function <- function(x) {
  return(0.5 * exp(-abs(x)))
}
```

The function for Metropolis-Hastings algorithm is defined that return an array of samples generated from the PDF distribution using the algorithm when called.
```{r}
metropolis_hastings <- function(initial_value, N, s) {
  samples <- numeric(N)
  samples[1] <- initial_value
  
  for (i in 2:N) {
    x_minus_1 <- samples[i - 1]
    
    x_random <- rnorm(1, mean = x_minus_1, sd = s) #a random number will be generated from a normal distribution centred around the previous number x_minus_1
    
    ratio <- pdf_function(x_random) / pdf_function(x_minus_1) #formula to compute the ratio
    
    u <- runif(1) #generate a random number from a uniform distribution ranging from 0 to 1
    
    if (u < ratio) {
      samples[i] <- x_random
    } else {
      samples[i] <- x_minus_1 #if else statement that appends the chosen sample 
    }
  }
  
  return(samples) #return an array of samples after the iteration ends
}
```

Defining the parameters and calling the function to generate an array of samples.
```{r}
N <- 100000 #set parameters
s <- 1
initial_value <- 0

samples <- metropolis_hastings(initial_value, N, s)
```

Using the results a histogram and KDE plot is generated.
```{r}
samples_df <- data.frame(sample = samples)

x_values <- seq(-10, 10, length.out = 10000)
true_density <- pdf_function(x_values) #computing the true density values corresponding to each point of x_values.

# Create histogram and density plot
p1 <- ggplot(samples_df, aes(x = sample)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50, fill = "skyblue", alpha = 0.5) +
  geom_density(aes(color = "Sampled Distribution"), linewidth = 1) +
  ggtitle("Monte Carlo Markov Chain - Metropolis-Hastings") +
  xlab("x") +
  ylab("Density") +
  theme_minimal() +
  geom_line(data = data.frame(x = x_values, y = true_density), 
            aes(x, y, color = "True Distribution"), linetype = "dashed") +
  scale_color_manual(values = c("blue", "red"), 
                     labels = c("Sampled Distribution", "True Distribution")) +
  annotate("text", x = -7, y = 0.4, label = paste("Sample Mean:", round(sample_mean, 3)), color = "blue") +
  annotate("text", x = -7, y = 0.35, label = paste("Sample Std Dev:", round(sample_std, 3)), color = "blue")

print(p1)
```
We extract and print the mean and standard deviation of the samples.
```{r}
# Calculate sample mean and sample standard deviation
sample_mean <- mean(samples)
sample_std <- sd(samples)

cat("Sample Mean is ", sample_mean, "\n")
cat("Sample Standard Deviation is ", sample_std, "\n")
```

The function to compute R hat value is defined.
```{r}
calculate_r <- function(B, W) {
  return(sqrt((B + W) / W))
}
```

We define the parameters to address part(b) and create an array to store the r-values
```{r}
N <- 2000 #set parameters
J <- 4
s_values <- seq(0.001, 1, length.out = 1000)
R_Values <- numeric(length(s_values)) 
```

Random initial values are defined depending on the number of chain.
```{r}
initial_values <- sample(30, J, replace = FALSE)
```

Iterating over a sequence of s values and returning an array of r-values.
```{r}
for (i in seq_along(s_values)) {
  s <- s_values[i]
  
  chains <- lapply(initial_values, function(initial_value) {
    metropolis_hastings(initial_value, N, s) #for each iteration, the algorithm generate J chains for each s value using the algorithm,
  })
  
  chain_sample_means <- sapply(chains, mean) # computing within-chain means and variances
  chain_sample_variances <- sapply(chains, var)
  
  W <- mean(chain_sample_variances)  # computing W and M
  M <- mean(chain_sample_means)
  
  # Compute B
  B <- mean((chain_sample_means - M)^2) 
  
  # Calculate R
  R <- calculate_r(B, W)
  R_Values[i] <- R
}
```

Dataframe is created using the resulting R-values. The data frame will then be used to plot the graph to visualize values of R over a grid of s values.
```{r}
data <- data.frame(s_values = s_values, R_Values = R_Values)

# Plot the graph
ggplot(data, aes(x = s_values, y = R_Values)) +
  geom_line(color = "blue") +
  labs(x = "s values", y = "R values", title = "R values over a grid of s values interval between 0.001 and 1.") +
  theme_minimal()
```
The R-value when s = 0.001 is computed.
```{r}
# Filter the data for s = 0.001
r_hat_s001 <- data$R_Values[data$s_values == 0.001]

print(paste("R hat value for s = 0.001:", r_hat_s001))
```
